# Data frame modification (general modification)
For the most of the time, we will have to engineer our data sets before we do some data analyzation on them. Here I present some tricks to modify our dat sets in R.

## 1.Create DataFrame by arrays
Creating DataFrame by two arrays, These two arrays will by columns of df.Array length should be the same.order matters.
```{r}
first_column <- c("value_1", "value_2", "value_3")
second_column <- c("value_4", "value_5", "value_6")
third_column <- c("value_7", "value_8", "value_9")
df <- data.frame(first_column, second_column)
df
```
another way to create dataFrame, both method produce the same result.
```{r}
df <- data.frame (first_column  = c("value_1", "value_2", "value_3"),
                  second_column = c("value_4", "value_5", "value_6"),
                  third_column = c("value_7", "value_8", "value_9")
                  )
df
```
## 2. Create DataFrame by combining two dataFrame(using cbind/rbind)
```{r}
# create a new dataFrame
df1 <- data.frame (first_column  = c("value_1", "value_2", "value_3"),
                  second_column = c("value_4", "value_5", "value_6"),
                  third_column = c("value_7", "value_8", "value_9")
                  )
df2 <- data.frame (third_column  = c("value_7", "value_8", "value_9"),
                  fourth_column = c("value_10", "value_11", "value_12")
                  )
# combine two dataframe by columns
# order matters.
bigger_df <- cbind(df1, df2)
print(bigger_df)
# combine two dataframe by rows
# two DataFrames must have the same column name and order.
df3 <- data.frame (third_column = c("value_7", "value_8", "value_9"),
                  fourth_column = c("value_10", "value_11", "value_12")
                  )
bigger_df2 <- rbind(df2,df3)
print(bigger_df2)
# order matters.
```
## 3. adding/change row name
```{r}
print(bigger_df2)
rownames(bigger_df2) <- LETTERS[16:21]

bigger_df2 <- data.frame(bigger_df2,
                   row.names = LETTERS[16:21])
# both ways produce the same result
print(bigger_df2)
# rownames can change the row name.
rownames(bigger_df2) <- 1:6
print(bigger_df2)
## get row names
rownames(bigger_df2)
```

## 4. changing column names
```{r}
print(df)
# change all column names
colnames(df) <- c('C1','C2')
print(df)
# change specific column name
colnames(df)[1] <- c('new column name')
print(df)
#get column names
colnames(df)
```

## 5. Adding array to to dataFrame as new column
```{r}
print(bigger_df2)
bigger_df_new = bigger_df2

bigger_df_new$new <- c(3, 3, 6, 7, 8, 12)

bigger_df_new['new'] <- c(3, 3, 6, 7, 8, 12)

new <- c(3, 3, 6, 7, 8, 12)
bigger_df_new <- cbind(df, new)
# three ways produce the same result.
print(bigger_df_new)
```

## 6. adding column by combination of other columns
```{r}
# using mutate() from dplyr
library(dplyr)
df_math <- data.frame (first_column  = c(1, 2, 3),
                  second_column = c(4, 5, 6)
                  )

## add new column as mathematical operation of other columns.
output <- mutate(df_math,
                 sum = (first_column + second_column) / 2)
print(output)
## add new column by boolean operation of other columns.
output <- mutate(df_math,
                 divisible_by2 = case_when(
                   first_column%%2 == 0 ~ "yes",
                   TRUE ~ "No" # otherwise not divisiable by 2.
                 ))
print(output)
```

## 7. joining two table **[1]**
```{r}
emp_df=data.frame(
  emp_id=c(1,2,3,4,5,6),
  name=c("Smith","Rose","Williams","Jones","Brown","Brown"),
  superior_emp_id=c(-1,1,1,2,2,2),
  dept_id=c(10,20,10,10,40,50),
  dept_branch_id= c(101,102,101,101,104,105)
)

dept_df=data.frame(
  dept_id=c(10,20,30,40),
  dept_name=c("Finance","Marketing","Sales","IT"),
  dept_branch_id= c(101,102,103,104)
)
# inner join
print(merge(x = emp_df, y = dept_df, by = "dept_id"))
# outer join 
print(merge(x = emp_df, y = dept_df, by = "dept_id", all = TRUE)) # all means containing all rows
# left join
print(merge(x = emp_df, y = dept_df, by = "dept_id", all.x = TRUE)) # all.x means containing all rows in x
# right join
print(merge(x = emp_df, y = dept_df, by = "dept_id", all.y = TRUE)) # all.y means containing all rows in y

```

## 8. pivoting dataframe, understanding pivot_longer
```{r}
# pivot_longer transform a data frame from a wide format to a long format by converting feature names to a categorical feature.

library(tidyr) # pivot_longer is from the package tidyr.
needs_pivoting <- data.frame(sticker_type=c('A', 'B', 'C', 'D'),
                 sparrow=c(12, 15, 19, 19),
                 eagle=c(22, 29, 18, 12))

print(needs_pivoting)

needs_pivoting %>% pivot_longer(cols=c('sparrow', 'eagle'), # the columns(feature names) to be pivoted
                                names_to='species', # the name of the column of features
                                values_to='price') # the name of the column of values
```

### 9. binning using cut()
```{r}
hour_df <- data.frame(shop_name=c('MAC', 'Tangro', 'cummington', 'Burger King', 'judgement', 'KFC', 'ye', 'Dungeon', 'Razer', 'yeah sir', 'Koban wife', 'string'),
                 operating_hours=c(2, 5, 4, 7, 7, 8, 5, 4, 5, 11, 13, 8),
                 rebounds=c(7, 7, 4, 6, 3, 8, 9, 9, 12, 11, 8, 9))
hour_df
new_hour_df <- hour_df %>% mutate(operating_hours_bin = cut(operating_hours, breaks=c(0,5,9,13)))
new_hour_df
```

### 10. binning using case()
```{r}
hour_df2 <- data.frame(shop_name=c('MAC', 'Tangro', 'cummington', 'Burger King', 'judgement', 'KFC', 'ye', 'Dungeon', 'Razer', 'yeah sir', 'Koban wife', 'string'),
                 operating_hours=c(2, 5, 4, 7, 7, 8, 5, 4, 5, 11, 13, 8),
                 rebounds=c(7, 7, 4, 6, 3, 8, 9, 9, 12, 11, 8, 9))
hour_df
new_hour_df <- hour_df2 %>% mutate(operating_hours_bin = case_when( # logistics
                   operating_hours <= 3 ~ 'very short',
                   operating_hours <= 6 & operating_hours > 3 ~ 'short',
                   operating_hours <= 10 & operating_hours > 6 ~ 'median',
                   operating_hours > 10 ~ 'long',
                   TRUE ~ 'what else will this be?'
                 ))
new_hour_df
```

## 11. ordering
```{r}
student_result_wild=data.frame(name=c("Ram","Geeta","John","Paul",
                                 "Cassie","Geeta","Paul"),
                          maths=c(7,8,8,9,10,8,9),
                          science=c(5,7,6,8,9,7,8),
                          history=c(7,7,7,7,7,7,7),
                          id = c(9,2,3,5,13,2,5))
student_result_wild # data with out ordering
# order data by certain variable (ascending)
student_result_wild[order(student_result_wild$id),]
# order data by certain variable (descending)
student_result_wild[order(-student_result_wild$id),]
# order data by multiple variable, the second orders the duplicates in first variable
student_result_wild[order(-student_result_wild$id, student_result_wild$science),]
```

## reference
[1] https://sparkbyexamples.com/r-programming/how-to-do-left-join-in-r/#:~:text=How%20to%20do%20left%20join%20on%20data%20frames%20in%20R,join%20data%20frames%20in%20R.